---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## 1 Loading and preprocessing the data

 - Read in the dataset
 - Remove incomplete rows
 - Convert the Data field to a Date Object (Factor object -> "character string" -> Date object)
 - Setup the Figures directory
 - Review the data:

```{r}
        myData <- read.csv("activity.csv", na.strings = "NA", header = TRUE)
        myData <- subset(myData, steps != "NA")  ## remove rows with "Not Available" data
        myData <- as.data.frame(lapply(myData,function (y) if(class(y)=="factor" ) as.character(y) else y),stringsAsFactors=F)
        myData$date <- as.Date(substr(myData$date,1,10), "%Y-%m-%d") ## Convert character Date to Date object
        knitr::opts_chunk$set(fig.path='Figures/')
        strMyData <- str(myData)
```

`r strMyData`

## 2 What is mean total number of steps taken per day?

```{r}
        byDay<-split(myData,myData$date, drop=TRUE)
        byDay <- sapply(byDay, function(x) sum(x$steps))  

        byDayMean <- aggregate(myData$steps, by=list(myData$date), FUN=mean)
        names(byDayMean) <- c("Date", "AverageNoSteps")

        myTitle <- paste("Steps per Day Removing NAs (mean=", round(mean(byDay, na.rm = TRUE), digits = 0), ")" )
        histInfo <- (hist(byDay, breaks=25, main = myTitle, xlab = "Steps", xlim=c(0,24000), ylim=c(0, 12)))
        abline(v = mean(byDay, na.rm = TRUE), col = "blue", lwd = 2)
        summary(byDay)
```

## 3 What is the average daily activity pattern?

```{r}
        byInterval <- aggregate(myData$steps, by=list(myData$interval), FUN=sum)
        names(byInterval) <- c("Interval", "AverageNoSteps")
        index <- with(byInterval, order(byInterval$AverageNoSteps, decreasing = TRUE))
        myTitle <- "Daily Activity Pattern"
        plot(byInterval$Interval, byInterval$AverageNoSteps, main = myTitle, type="l", ylim=c(0,12000), pch="|", xlab = "Time Interval", ylab = "Steps", xaxt="n")
        axis(1, at = seq(0, 2400, by = 100), las=2)
        abline(v = byInterval[index,][1,1], col = "blue", lwd = 2 )
        myLineAnnotation <- paste("Max Avg Steps =", round(byInterval[index,][1,2], digits=0), " @ Interval =", byInterval[index,][1,1])
        text(1550, 11000, myLineAnnotation)
        summary(byInterval)
```

## 4 Inputing missing values

```{r}
        myData2 <- read.csv("activity.csv", na.strings = "NA", header = TRUE)
        myData2 <- as.data.frame(lapply(myData2,function (y) if(class(y)=="factor" ) as.character(y) else y),stringsAsFactors=F)
        myData2$date <- as.Date(substr(myData2$date,1,10), "%Y-%m-%d")
        
        rowTotal1 <- nrow(myData2)
        rowComplete1 <- as.integer(sum(as.numeric(complete.cases(myData2))))
        rowDiff1 <- rowTotal1-rowComplete1
        rowStepsWithNA1 <- sum(as.numeric(is.na(myData2$steps)))
        
        for(i in 1:nrow(myData2)){
                if (is.na(myData2$steps[i])) {
                        if (length(byDayMean$AverageNoSteps[byDayMean$Date == myData2$date[i]]) != 0) {
                                myData2$steps[i] <- byDayMean$AverageNoSteps[byDay$Date == myData2$date[i]]
                        }
                        else { myData2$steps[i] = 0 }                        
                }
        }
        
        rowTotal2 <- nrow(myData2)
        rowComplete2 <- as.integer(sum(as.numeric(complete.cases(myData2))))
        rowDiff2 <- rowTotal2-rowComplete2
        rowStepsWithNA2 <- sum(as.numeric(is.na(myData2$steps)))
        
        byDayAfter<-split(myData,myData$date, drop=TRUE)
        byDayAfter <- sapply(byDayAfter, function(x) sum(x$steps))  
        myTitle <- paste("Steps per Day Adjusting NA Data (mean=", round(mean(byDayAfter), digits = 0), ")" )
        histInfo <- (hist(byDayAfter, breaks=25, main = myTitle, xlab = "Steps", xlim=c(0,24000), ylim=c(0, 12)))
        abline(v = mean(byDayAfter, na.rm = TRUE), col = "blue", lwd = 2)
        summary(byDayAfter) 
```

Data statistics *before* adjusting for missing values (ignoring missing values) - 
        
        - Number of incomplete rows:  total number of rows - number of complete rows = `r rowTotal1` - `r rowComplete1` = `r rowDiff1`  
        - Number of rows with 'steps == NA':  `r rowStepsWithNA1`
        - Average steps :  `r summary(byDay)[4]` 
        - Median steps :  `r summary(byDay)[3]` 
        
        Data statistics *after* adjusting missing values - 
                
                - Number of incomplete rows:  total number of rows - number of complete rows = `r rowTotal2` - `r rowComplete2` = `r rowDiff2`  
        - Number of rows with 'steps == NA':  `r rowStepsWithNA2`
        - Average steps :  `r summary(byDayAfter)[4]`  
        - Median steps :  `r summary(byDayAfter)[3]` 
        
The mean and median are the same for both datasets.

## 5 Differences in activity patterns between weekdays and weekends

```{r}
        myData2$timeframe <- ifelse((weekdays(myData2$date) == c("Saturday")) | (weekdays(myData2$date) == c("Sunday")),"weekend", "weekday")
        myData2$timeframe <- factor(myData2$timeframe)
        
        byInterval <- aggregate(myData2$steps, by=list(myData2$interval, myData2$timeframe), FUN=sum)
        names(byInterval) <- c("Interval", "timeframe", "AverageNoSteps")
        index <- with(byInterval, order(byInterval$AverageNoSteps, decreasing = TRUE))
        
        library(ggplot2)
        ggplot(byInterval, aes(byInterval$Interval, byInterval$AverageNoSteps)) +
                           xlab("Interval") +
                           ylab("Number of Steps") +
                           geom_line() +
                           scale_x_continuous(breaks = seq(0, 2400, by=100)) +
                           facet_wrap(~timeframe, nrow=2) + 
                           theme(axis.text.x = element_text(angle = 90, hjust = 1))
        
        summary(byInterval)
```
